---
title: "Week 6 - Code Along Lab"
author: "Hannah Brady"
date: "11/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(ggpubr)
library(broom)
```

# Week 6 Coding Lesson - Intro to Linear Regression

The objectives of Lab Week 6 are:

+ Example of rank-based tests for rank / medians comparison (Mann-Whitney U)
+ Simple linear regression by OLS in R with `lm()`
+ Check assumptions of OLS (diagnostic plots with `plot()`)
+ Visualize linear model, and summarize in text
+ Find correlation by Pearson’s *r*

## 1. Attached packages:

+ `library(tidyverse)`
+ `library(palmerpenguins)`
+ `library(ggpubr)`
+ `library(broom)`

## 2. A rank-based test example (Mann Whitney U)

Recall the following two, non-parametric, rank-based hypothesis tests we learned this week:

+ Mann-Whitney U to compare ranks (medians) between two unpaired samples (non-parametric alternative to two-sample t-test)
+ Kruskall-Wallis to compare ranks (medians) between > 2 samples (non-parametric alternative to one-way ANOVA)

As an example we will make mock unpaired data and investigate differences in ranks, "medians comparison," by Mann-Whitney U using the `wilcox.test()` function.

Note: You’d also use this for a paired Wilcoxon-Signed-Rank test, with an additional ‘paired = TRUE’ argument, if samples were paired.

This code chunk makes two sample vectors `gp_1` and `gp_2` that are the same as Allison's random samples because we use `set.seed()`. `sample.int` creates random sample integers.

```{r}
set.seed(1414)
gp_1 <- sample.int(20, size = 15, replace = TRUE)
gp_1

set.seed(1424)
gp_2 <- sample.int(30, size = 15, replace = TRUE)
gp_2
```

Create 2 exploratory histograms:

```{r}
hist(gp_1)
hist(gp_2)
```

Histogram interpretation:

1. Not clearly normally distributed from exploratory histograms
2. Somewhat small sample size (n = 15 for each)
3. sometimes: I’ve decided that ranks (or, medians) are a more valuable metric to compare for these data.

So do a rank-based test! We will do Mann Whitney U using `wilcox.test()`

```{r}
my_mwu <- wilcox.test(gp_1, gp_2)

# Note: you will get a warning here that is just a "heads up"
# If there are ties in ranks, the p-value is estimated using a normal approximation (and is fine)

my_mwu
```

p-value interpretation:

If the null hypothesis is true (these samples were drawn from populations with the same median), there is a probability of 0.28 that we could have found median values *at least as different* as ours by chance. In other words: not sufficient evidence to reject the null hypothesis of equal ranks (or medians) using a significance level of 0.05.

Though not doing it today, see `?kruskal.test` in console for more information about a rank-based test for comparing medians across > 2 groups (i.e. the rank-based alternative to one-way ANOVA).

## 3. Simple linear regression

For this example we will compare penguin flipper length versus body mass from the `palmerpenguins` package.

### Exploratory viz

Start with exploratory data viz (note in the future we probably want to thnk of species and sex separately, too):

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

We should ask questions about our exploratory visualization, like:

+ Does it look like a linear relationship makes sense?
+ Do we have any concerns about modeling as a linear relationship?
+ Any notable outliers?
+ Initial thoughts about homoscedasticity (explored more later)?

Here, it looks like overall a linear relationship between flipper length and body mass makes sense here (moving forward, we’re learn how to include species and sex as part of the model, but for now we’ll just use the single exploratory variable flipper_length_mm).

### Model (linear regression)

Once we’ve decided that a linear relationship makes sense, we’ll model it using `lm()`.

Note that we haven’t checked all assumptions yet. That’s because a lot of our assumptions for linear regression are based on model residuals (e.g. normality & homoscedasticity of residuals), which we can’t calculate until after we find the predicted values from the model.

Recall that the $residual = y_{actual} - y_{predicted}$

```{r}
# Linear model with body mass dependent on flipper length.
penguin_lm <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
summary(penguin_lm)
```

Check the Week 6 Part 2 lecture video for information about how to interpret different pieces of this output, but a couple to highlight:

+ Pr(>|t|) tells us both the intercept and flipper_length_mm coefficients are significantly different from zero (not super interesting)
+ The Multiple R^2^ value is 0.759 - meaning that 75.9% of variance in body mass is explained by flipper length

### Access model outputs

Use the `broom::tidy()` function to get the model outputs in nice data frame format:

```{r}
penguin_lm_tidy <- broom::tidy(penguin_lm)
penguin_lm_tidy
```

We want to save some of the outputs so we can easily reference them in-text and have them automatically update if the model is updated:

```{r}
# Intercept:
penguin_int <- penguin_lm_tidy$estimate[1]
penguin_int

# Coefficient:
penguin_coef <- penguin_lm_tidy$estimate[2]
penguin_coef
```

Use `broom::glance()` to get other model information (degrees of freedom, F-statistic, p-value, etc.):

```{r}
# Metrics at a glance: 
penguin_lm_out <- broom::glance(penguin_lm)
penguin_lm_out
```

Writing a statement with these values that automatically updates if the model is changed:

"Simple linear regression was used to explore the relationship between penguin flipper length (mm) and body mass (g) across all three penguin species, and including both male and female penguins. A significant regression model was found ($\beta$ = `r round(penguin_coef,3)`, F(`r penguin_lm_out$df`,`r penguin_lm_out$df.residual`) = `r round(penguin_lm_out$statistic,1)`, p < 0.001) with an R^2^ of `r round(penguin_lm_out$r.squared,3)`."

Note that Allison uses “p < 0.001” here if the p-value is very small - this is somewhat standard.

### Explore model assumptions

Recall that we have assumptions for linear regression we need to explore, some related to the residuals.

+ Linearly related variables (CHECK - already looked & thought hard)
+ Normally distributed residuals
+ Homoscedasticity (constant residuals variance)
+ iid residuals (no serial correlation) - more often a concern in time series data

Use the `plot()` function on the model, which will automatically create four useful visualizations to consider assumptions!

```{r}
plot(penguin_lm)
```

Notice that four plots show up. What do they show? Make sure to watch Part 2 of the lecture, which discusses how we can interpret each of these diagnostic plots. 

- **The first one**: fitted values vs. residuals 
- **The second one**: QQ-plot for residuals 
- **The third one**: another way of looking at fitted vs. residuals (these are just standardized residuals, but you can interpret it the same way)
- **The fourth one**: Cook's distance, a measure of "influence" or "leverage" that individual points have on the model - often considered a way to explore outliers. 

See the Week 6 Part 2 Lecture video for more information about how to interpret these outcomes, but in summary: graphs 1 & 3 are useful for thinking about homoscedasticity; graph 2 (QQ plot) helps us consider normality of residuals; graph 4 reveals the Cook's distance (a measure of how much leverage any single observation has on the model).

### Visualize the model

Now that we've explore the assumptions and have decided that linear regression is a valid tool to describe the relationship between flipper length and body mass, let's look at the model.

- Use `geom_smooth(method = "lm")` to add a linear model to an existing scatterplot

- Use `stat_cor()` and/or `stat_regline_equation()` to add equation information directly to the plot panel, at an x- and y-position that you specify (and yes, you can mess with the digits & appearance here)

```{r}

ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm",
              color = "red",
              size = 0.5,
              fill = "gray10",
              alpha = 0.5) +
  theme_light() +
  ggpubr::stat_regline_equation(label.x = 180, label.y = 5700)
  
```

### Find Pearson's *r* for correlation: 

In lecture we talked about the coefficient of determination, R^2^, which tells us how much of the variance in the dependent variable is explained by the model. 

We might also want to explore the strength of the correlation (degree of relationship) between two variables which, for two linearly related continuous variables, can be expressed using Pearson's *r*. 

Pearson's *r* ranges in value from -1 (perfectly negatively correlated - as one variable increases the other decreases) to 1 (perfectly positively correlated - as one variable increases the other increases). A correlation of 0 means that there is no degree of relationship between the two variables. 

Typical guidelines look something like this (there's wiggle room in there): 

- *r* = 0: no correlation
- *r* < |0.3|: weak correlation
- *r* between |0.3| and |0.7|: moderate correlation
- *r* > |0.7|: strong correlation

We'll use the `cor.test()` function, adding the two vectors (`flipper_length_mm` and `body_mass_g`) as the arguments. The function reports the Pearson's *r* value, and performs a hypothesis test with null hypothesis that the correlation = 0. 

```{r}
penguins_cor <- cor.test(penguins$flipper_length_mm, penguins$body_mass_g)
penguins_cor
```

Here, we see that there is a strong positive correlation between penguin flipper length and body mass (*r* = `r round(penguins_cor$estimate,2)`, t(`r penguins_cor$parameter`) = `r round(penguins_cor$statistic,2)`, p < 0.001). 

**Note**: Once you have a "template" statement, you can just replace `penguins_cor` here with whatever your correlation analysis is stored as! You don't need to recreate the wheel every time! 

## END LAB - CONGRATULATIONS!
